---
title: "optimal_difficulty_lmer_analysis"
author: "Erva"
date: "2024-04-26"
output: html_document
---

Packages

```{r echo=FALSE}
library(data.table)
library(ggplot2)
library(dplyr)
library(tools)
library(progress)
library(viridis)
library(ggforce)
library(see)
library(stringr)
library(reshape2)
library(gridExtra)
library(lme4)
library(lmerTest)
library(sjPlot)

```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/Ghislane/Desktop/optimal_difficulty_1/code/data/sides/elo_bins')
```

# read file
```{r}
file= "average_quest_dif_100.csv"
# read the file
average_quest_dif <- read.csv(file)

  name_file <- substr(file, 18, nchar(file))

```

# rename variables to make them transparent
```{r}
# rename the column
average_quest_dif <- average_quest_dif %>%
  rename(mean_relative_difficulty = mean_skill_difficulty_difference)

# rename the column
average_quest_dif <- average_quest_dif %>%
  rename(student_ability = ability_2020_2021)

# rename the column
average_quest_dif <- average_quest_dif %>%
  rename(relative_difficulty_slope = slope)
```


# mean- center interaction variables  (only the n_question_in_spec_training)
```{r}
# Calculate the mean of n_question_in_spec_training
mean_n_question <- mean(average_quest_dif$n_question_in_spec_training, na.rm = TRUE)

# Center the variable n_question_in_spec_training by subtracting its mean from each observation
average_quest_dif$n_question_in_spec_training_c <- average_quest_dif$n_question_in_spec_training - mean_n_question

```

# correaltions
## correlation matrix
```{r}

# Selecting the columns
for_table <- average_quest_dif[c('mean_relative_difficulty', 'student_ability', 'prop_correct_ecn', 'mean_student_average_ability', 'n_question_in_spec_training', 'mean_difficulty','mean_elo_ExpectedScore','n_question_in_spec_training_c')]


# Calculating means and standard deviations
means <- colMeans(for_table, na.rm = TRUE)
sds <- apply(for_table, 2, sd, na.rm = TRUE)

# Correlation matrix
correlations <- cor(for_table, use = 'pairwise.complete.obs')

# Create a formatted table with means, SDs, and correlations
formatted_table <- data.frame(M = means, SD = sds, correlations)

# Function to determine significance asterisks
significance_asterisks <- function(p_value) {
  if (p_value < 0.001) {
    return('***')
  } else if (p_value < 0.01) {
    return('**')
  } else if (p_value < 0.05) {
    return('*')
  } else {
    return('')
  }
}

# Populate the correlations with significance testing
for (col in colnames(for_table)) {
  for (row in colnames(for_table)) {
    if (col == row) {
      # Diagonals are not displayed in the table
      formatted_table[row, col] <- '—'
    } else {
      # Calculate the p-value
      p_value <- Hmisc::rcorr(as.matrix(for_table[, c(row, col)]))$P[1, 2]
      # Format with two decimal places and add significance asterisks
      corr <- correlations[row, col]
      formatted_table[row, col] <- paste0(format(corr, digits = 2), significance_asterisks(p_value))
    }
  }
}

# You can save this table to a CSV or Excel file, or print it out
write.csv(formatted_table, file = 'formatted_correlation_table_with_significance.csv', row.names = TRUE)

formatted_table


# Calculating the correlation matrix
correlation_matrix <- cor(for_table, use = "pairwise.complete.obs")

# Print the correlation matrix
print(correlation_matrix)


# visiulazi
# Compute correlation matrix
correlation_matrix <- cor(for_table)

# Visualize the correlation matrix
library(corrplot)
corrplot(correlation_matrix, method = "number")


```


## within student correlation
```{r}
library(dplyr)

# Calculate the correlation within each student across their specialties
student_correlations <- average_quest_dif %>%
  group_by(student) %>%
  summarize(correlation = cor(mean_relative_difficulty, student_ability, use = "complete.obs"),
            .groups = 'drop')

# Viewing the summary of these correlations
summary(student_correlations$correlation)

# Plotting the distribution of within-student correlations
p1 <- ggplot(student_correlations, aes(x = correlation)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.05, fill = "blue", alpha = 0.3, color = "black") +
  geom_density(color = "blue", fill = "blue", alpha = 0.5) +
  ggtitle("Density and Histogram of Within-Student Correlations") +
  xlab("Correlation between Mean Relative Difficulty and Ability") +
  ylab("Density") +
  theme_minimal()

print(p1)

```


## within specialty correlation

```{r}
library(dplyr)

# Calculate the correlation within each student across their specialties
specialty_correlations <- average_quest_dif %>%
  group_by(specialty) %>%
  summarize(correlation = cor(mean_relative_difficulty, student_ability, use = "complete.obs"),
            .groups = 'drop')

print(specialty_correlations)
# Viewing the summary of these correlations
summary(specialty_correlations$correlation)

# Plotting the distribution of within-specialty correlations using a bar plot
p1 <- ggplot(specialty_correlations, aes(x = specialty, y = correlation)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.3, color = "black") +
  ggtitle("Bar Plot of Within-specialty Correlations") +
  xlab("Specialty") +
  ylab("Correlation between Mean Relative Difficulty and Ability") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

print(p1)



```


### Full Model & Colinearity check : specialty as radnom intercept and slope
```{r}
model_formula <- prop_correct_ecn ~ I(mean_relative_difficulty^2) + mean_relative_difficulty + 
  n_question_in_spec_training_c+
  student_ability+
  student_ability:I(mean_relative_difficulty^2)+
  student_ability:mean_relative_difficulty+
  student_ability:n_question_in_spec_training_c+
  (1 | student)+
  (1 + mean_relative_difficulty + I(mean_relative_difficulty^2) | specialty)

# Fit the linear mixed-effects model
model_fit <- lmer(model_formula, data = average_quest_dif, control = lmerControl(optimizer = "bobyqa"))

# Display the model summary with the name of the file
cat(paste("Model summary for", name_file, ":\n"))
print(summary(model_fit))
# Create a table of fixed effects
tabel_model <- tab_model(model_fit, show.se = TRUE, title = "Table: Regression Analysis Results")

# Print the table
tabel_model

## CHECK COLINEARITY
library(performance)
check_collinearity(model_fit)


## AIC and BIC - Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC)
aic_value <- AIC(model_fit)
bic_value <- BIC(model_fit)
cat("AIC =", aic_value, "\n")
cat("BIC =", bic_value, "\n")

```


### p value for random effect
```{r}
library(lme4)
full_model <- lmer(prop_correct_ecn ~ I(mean_relative_difficulty^2) + mean_relative_difficulty + 
  n_question_in_spec_training_c +
  student_ability +
  student_ability:I(mean_relative_difficulty^2) +
  student_ability:mean_relative_difficulty +
  student_ability:n_question_in_spec_training_c +
  (1 | student) +
  (1 + mean_relative_difficulty + I(mean_relative_difficulty^2) | specialty), 
  data = average_quest_dif, control = lmerControl(optimizer = "bobyqa"))


reduced_model <- lmer(prop_correct_ecn ~ I(mean_relative_difficulty^2) + mean_relative_difficulty + 
  n_question_in_spec_training_c +
  student_ability +
  student_ability:I(mean_relative_difficulty^2) +
  student_ability:mean_relative_difficulty +
  student_ability:n_question_in_spec_training_c +
  (1 | student) +
  (1 + I(mean_relative_difficulty^2) | specialty), 
  data = average_quest_dif, control = lmerControl(optimizer = "bobyqa"))

library(lmtest)
anova_result <- anova(reduced_model, full_model)
print(anova_result)

```


## correlation fixed effects
```{r}
# Convert fixed effects coefficients to a matrix-like object
fixed_effects_matrix <- matrix(fixed_effects, nrow = 1)

# Calculate correlation matrix for fixed effects
cor_fixed_effects <- cor(t(fixed_effects_matrix))

# Display the correlation matrix
cat("Correlation matrix of fixed effects:\n")
print(cor_fixed_effects)

```


##have table with exact numbers

```{r}
# Assume we already have the summary data frame
fixed_effects <- summary(model_fit)$coefficients

# Check if p-values are below .Machine$double.eps and adjust display
fixed_effects[,"Pr(>|t|)"] <- ifelse(fixed_effects[,"Pr(>|t|)"] < .Machine$double.eps,
                                     "< 1e-16",  # You can choose the threshold
                                     format.pval(fixed_effects[,"Pr(>|t|)"], digits = 10))

# Print the modified fixed effects table
print(fixed_effects)

```


# Model Fit
```{r}
## PLOT
  
  # Create a sequence of mean_relative_difficulty values for plotting
  x_values <- seq(
    min(average_quest_dif$mean_relative_difficulty ),
    max(average_quest_dif$mean_relative_difficulty ),
    length.out = 20
  )

  x_values_slope <- seq(
    min(average_quest_dif$relative_difficulty_slope ),
    max(average_quest_dif$relative_difficulty_slope ),
    length.out = 20
  )
  
   x_values_ability <- seq(
    min(average_quest_dif$student_ability ),
    max(average_quest_dif$student_ability ),
    length.out = 20
  )
   
   x_values_nb_question<-seq(
    min(average_quest_dif$n_question_in_spec_training_c ),
    max(average_quest_dif$n_question_in_spec_training_c ),
    length.out = 10
  )
   
  # List of unique specialties
  specialties <- unique(average_quest_dif$specialty)
  
  
  # Create a data frame to store the values for prediction
prediction_data <- expand.grid(
  mean_relative_difficulty = x_values,
  specialty = unique(average_quest_dif$specialty),
  student_ability = x_values_ability,
  n_question_in_spec_training_c=mean(average_quest_dif$n_question_in_spec_training_c ),
  relative_difficulty_slope=mean(average_quest_dif$relative_difficulty_slope )
)
  
  # Predict values using the model, including random effects for 'specialty'
  predicted_values <- predict(model_fit, newdata = prediction_data, re.form = ~   (1 + mean_relative_difficulty + I(mean_relative_difficulty^2) | specialty))
  
  prediction_data$learning_slope = predicted_values
  
  # Convert student to factor
  average_quest_dif <- average_quest_dif %>%
    mutate(student = factor(student))
  
  library(ggplot2)
  # 
  # # find the mean of mean relative diffciulty
  # mean_mean_relative_diffciulty= prediction_data %>%
  # group_by(specialty) %>%
  #    summarise(mean_mean = mean(mean_relative_difficulty))

# Find the mean_relative_difficulty values that maximize the predicted prop_correct_ecn for each specialty and student_ability
max_x_values <- prediction_data %>%
  group_by(specialty, student_ability) %>%
  summarise(max_mean_difficulty = mean_relative_difficulty[which.max(learning_slope)])

  # Calculate the number of points in each facet for each specialty
facet_counts <- average_quest_dif %>%
  group_by(specialty) %>%
  summarise(n = n())
  
p <- ggplot(prediction_data, aes(x = mean_relative_difficulty, y = learning_slope)) +
  geom_line(aes(group = interaction(specialty, student_ability), color = student_ability),
            size = 2, alpha = 0.7) +
  labs(
    x = expression(atop("Mean Relative Difficulty", atop("Mean(Question Difficulty - Online Student Ability)", ""))),
    y = expression(atop("Final Exam Score", atop("(Proportion Correct)", "")))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 16, margin = margin(b = 10)),
    plot.caption = element_text(size = 14, margin = margin(t = 10)),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 11),
    panel.grid.major = element_line(color = "lightgray", linetype = "dashed")
  ) +
  facet_wrap(~gsub("_", " ", specialty), scales = "free", ncol = 4) +  # 4 columns
  scale_color_viridis_c(option = "C", direction = -1) +
  geom_vline(
    data = max_x_values,
    aes(xintercept = max_mean_difficulty, color = student_ability),
    size = 0.7,
    linetype = "dashed"
  ) +
  geom_text(
    data = facet_counts,
    aes(label = paste("n =", n), x = Inf, y = Inf),
    hjust = 1,
    vjust = 1,
    size = 5,
    show.legend = FALSE
  ) +
  labs(color = "Final Student Ability", linetype = "Optimal")

# Increase the width for 4 subplots per row
ggsave(
   "C:/Users/Ghislaine/Desktop/optimal_difficulty_1/code/data/sides/elo_bins/modelfit.png",
  plot = p,
  width = 19, height = 10, units = "in", dpi = 300
)

print(p)
```


## model fit for only students with mean student ability

```{r}
## PLOT
  
  # Create a sequence of mean_relative_difficulty values for plotting
  x_values <- seq(
    min(average_quest_dif$mean_relative_difficulty ),
    max(average_quest_dif$mean_relative_difficulty ),
    length.out = 20
  )

  x_values_slope <- seq(
    min(average_quest_dif$relative_difficulty_slope ),
    max(average_quest_dif$relative_difficulty_slope ),
    length.out = 20
  )
  
   x_values_ability <- seq(
    min(average_quest_dif$student_ability ),
    max(average_quest_dif$student_ability ),
    length.out = 20
  )
   
   x_values_nb_question<-seq(
    min(average_quest_dif$n_question_in_spec_training_c ),
    max(average_quest_dif$n_question_in_spec_training_c ),
    length.out = 10
  )
   
  # List of unique specialties
  specialties <- unique(average_quest_dif$specialty)
  
  
  # Create a data frame to store the values for prediction
prediction_data <- expand.grid(
  mean_relative_difficulty = x_values,
  specialty = unique(average_quest_dif$specialty),
  student_ability = mean(average_quest_dif$student_ability ),
  n_question_in_spec_training_c=mean(average_quest_dif$n_question_in_spec_training_c ),
  relative_difficulty_slope=mean(average_quest_dif$relative_difficulty_slope )
)
  
  # Predict values using the model, including random effects for 'specialty'
  predicted_values <- predict(model_fit, newdata = prediction_data, re.form = ~   (1 + mean_relative_difficulty + I(mean_relative_difficulty^2) | specialty))
  
  prediction_data$learning_slope = predicted_values
  
  # Convert student to factor
  average_quest_dif <- average_quest_dif %>%
    mutate(student = factor(student))
  
  library(ggplot2)
  # 
  # # find the mean of mean relative diffciulty
  # mean_mean_relative_diffciulty= prediction_data %>%
  # group_by(specialty) %>%
  #    summarise(mean_mean = mean(mean_relative_difficulty))

# Find the mean_relative_difficulty values that maximize the predicted prop_correct_ecn for each specialty and student_ability
max_x_values <- prediction_data %>%
  group_by(specialty, student_ability) %>%
  summarise(max_mean_difficulty = mean_relative_difficulty[which.max(learning_slope)])

  # Calculate the number of points in each facet for each specialty
facet_counts <- average_quest_dif %>%
  group_by(specialty) %>%
  summarise(n = n())
  
p <- ggplot(prediction_data, aes(x = mean_relative_difficulty, y = learning_slope)) +
  geom_line(aes(group = interaction(specialty, student_ability), color = student_ability),
            size = 2, alpha = 0.7) +
  labs(
    x = expression(atop("Mean Relative Difficulty", atop("Mean(Question Difficulty - Online Student Ability)", ""))),
    y = expression(atop("Mock Final Exam Score", atop("(Proportion Correct)", "")))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 16, margin = margin(b = 10)),
    plot.caption = element_text(size = 14, margin = margin(t = 10)),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16),
    strip.text = element_text(size = 11),
    panel.grid.major = element_line(color = "lightgray", linetype = "dashed")
  ) +
  facet_wrap(~gsub("_", " ", specialty), scales = "free", ncol = 4) +  # 4 columns
  scale_color_viridis_c(option = "C", direction = -1) +
  # geom_vline(
  #   data = max_x_values,
  #   aes(xintercept = max_mean_difficulty, color = student_ability),
  #   size = 1,
  #   linetype = "dashed"
  # ) +
  labs(color = "Final Student Ability", linetype = "Optimal")
  #+geom_text(data = max_x_values, aes(x = max_mean_difficulty-1.5, y = 0.3, label = round(max_mean_difficulty, 2)), vjust = -1, size = 7.5, color = "darkgreen", hjust = -0.1)


# Increase the width for 4 subplots per row
ggsave(
   "C:/Users/Ghislaine/Desktop/optimal_difficulty_1/code/data/sides/elo_bins/modelfit.png",
  plot = p,
  width = 19, height = 10, units = "in", dpi = 300
)

print(p)
```

# calculate the mean optimal relative difficulty across all specialties and students

```{r}
# Extract random effects
random_effects <- ranef(model_fit)

# Get the random effects for each specialty
specialty_effects <- random_effects$specialty

# Assuming a plausible range for mean_relative_difficulty
difficulty_range <- seq(-3, 3, length.out = 100)


library(dplyr)
library(purrr)

# Assume mean values for student_ability and n_question_in_spec_training_c
mean_student_ability <- mean(average_quest_dif$student_ability)
mean_n_questions <- mean(average_quest_dif$n_question_in_spec_training_c)

# Function to calculate predictions
calculate_predictions <- function(specialty) {
  # Extract specialty-specific random effects
  intercept_spec <- specialty_effects[[specialty]]$`(Intercept)`
  slope_difficulty <- specialty_effects[[specialty]]$mean_relative_difficulty
  slope_difficulty_sq <- specialty_effects[[specialty]]$`I(mean_relative_difficulty^2)`
  
  # Calculate predictions across the difficulty range
  sapply(difficulty_range, function(diff) {
    pred <- (fixed_effects$`(Intercept)`
             + fixed_effects$mean_relative_difficulty * diff
             + fixed_effects$`I(mean_relative_difficulty^2)` * diff^2
             + fixed_effects$student_ability * mean_student_ability
             + fixed_effects$`student_ability:mean_relative_difficulty` * mean_student_ability * diff
             + fixed_effects$`student_ability:I(mean_relative_difficulty^2)` * mean_student_ability * diff^2
             + fixed_effects$n_question_in_spec_training_c * mean_n_questions
             + fixed_effects$`student_ability:n_question_in_spec_training_c` * mean_student_ability * mean_n_questions
             + intercept_spec
             + slope_difficulty * diff
             + slope_difficulty_sq * diff^2)
    return(pred)
  })
}

# Extract fixed effects
fixed_effects <- fixef(model_fit)

fixed_effects <- as.list(fixef(model_fit))

# Calculate predictions for each specialty and find the maximum
optimal_difficulties <- lapply(rownames(specialty_effects), function(specialty) {
  predictions <- calculate_predictions(specialty)
  max_index <- which.max(predictions)
  max_difficulty <- difficulty_range[max_index]
  list(specialty = specialty, max_difficulty = max_difficulty, max_prediction = max(predictions))
})

optimal_difficulties <- bind_rows(optimal_difficulties)

print(optimal_difficulties)

```

```{r}
library(lme4)
library(dplyr)

# Assuming model_fit is already fitted

# Calculate mean values for student_ability and n_question_in_spec_training_c
mean_student_ability <- mean(average_quest_dif$student_ability, na.rm = TRUE)
mean_n_questions <- mean(average_quest_dif$n_question_in_spec_training_c, na.rm = TRUE)

# Define a sequence of mean_relative_difficulty values to evaluate
mrd_seq <- seq(from = min(average_quest_dif$mean_relative_difficulty, na.rm = TRUE), 
               to = max(average_quest_dif$mean_relative_difficulty, na.rm = TRUE), length.out = 100)

# Function to calculate predicted prop_correct_ecn for a given specialty
predict_specialty <- function(specialty) {
  # Create a data frame for predictions
  pred_data <- expand.grid(
    mean_relative_difficulty = mrd_seq,
    student_ability = mean_student_ability,
    n_question_in_spec_training_c = mean_n_questions,
    specialty = specialty
  )
  
  # Predict prop_correct_ecn using the model
  pred_data$prop_correct_ecn <- predict(model_fit, newdata = pred_data,  re.form = ~ (1 + mean_relative_difficulty + I(mean_relative_difficulty^2) | specialty))
  
  # Find the mean_relative_difficulty that maximizes prop_correct_ecn
  max_difficulty <- pred_data$mean_relative_difficulty[which.max(pred_data$prop_correct_ecn)]
  return(max_difficulty)
}

# Apply function to each specialty
specialties <- unique(average_quest_dif$specialty)
optimal_difficulties <- sapply(specialties, predict_specialty)

# View results
names(optimal_difficulties) <- specialties
optimal_difficulties

# Calculate mean and standard deviation of optimal mean_relative_difficulties
mean_difficulty <- mean(optimal_difficulties)
sd_difficulty <- sd(optimal_difficulties)

# Print results
cat("Mean of optimal mean_relative_difficulty: ", mean_difficulty, "\n")
cat("Standard deviation of optimal mean_relative_difficulty: ", sd_difficulty, "\n")



```

